---
title: "Practice with nMDS"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Sal's code for creating the dissimilarity matrix 
```{r}
library(tidyverse)
library(kernlab)
library(dbscan)
library(vegclust)
library(cluster)
library(ggsci)

path_of_code <- dirname(rstudioapi::getSourceEditorContext()$path)
setwd(path_of_code)

#Read in the code and start by analyzing ACSA2. The process used here can be applied automatically to the other species later on. 

df <- read_csv("clean_data/UMRS_FPF_clean.csv")
labels <- read_csv("clean_data/plot_classification.csv")
df_cols <- left_join(df, labels, by = "PID") %>% select(PID, TR_SP, BasalArea, TreesPerAcre, Type, Label)
df_acsa2 <- filter(df_cols, Type == "ACSA2")
#Only 28 species to deal with in the silver maple dominant communities!
length(unique(df_acsa2$TR_SP))

#TPA = 1 / (pi * (DBH*2.75)^2) / 43560)
#BA = 0.25*pi*(DBH^2)
TPA_bins = 1 / (pi * (seq(1:106)*2.75)^2) / 43560
BA_bins = 0.25*pi*(seq(1:106)^2)


#confirm TPA length
(1/sqrt(TPA_bins))[2] - (1/sqrt(TPA_bins))[1]
(1/sqrt(TPA_bins))[3] - (1/sqrt(TPA_bins))[2]

tpa_tf <- function(x){return(1/sqrt(x))}
ba_tf <- function(x){return(sqrt(x))}

#Confirm uniform distribution
qplot(1/sqrt(TPA_bins), binwidth = 1017.306)
qplot(sqrt(BA_bins), binwidth = sqrt(BA_bins)[2] - sqrt(BA_bins)[1])


#Stratify the vegetation data
test <- stratifyvegdata(df_acsa2, sizes1 = BA_bins, plotColumn = "PID", speciesColumn = "TR_SP", abundanceColumn = "TreesPerAcre", size1Column = "BasalArea", cumulative = TRUE )
#cap <- CAP(test)

#Plot for individual species in one plot as in De Caceres, 2013. Probably do not want to use
ggplot(filter(df_acsa2, PID == "Cuivre-1-41")) + geom_step(aes(x = BasalArea, y = TreesPerAcre, color = TR_SP )) + theme_light()

#Compare CAP for multiple plots
#example <- cap[[1]][rowSums(cap[[1]][,-1]) > 0,colSums(cap[[1]][-1,]) > 0]
#example

#plot(cap, plots = "1", sizes = BA_bins[1:5])

dissim <- vegdiststruct(test, method = "manhattan")
write_csv(as.data.frame(as.matrix(dissim)), "dissimilarity_matrix.csv")


```
```{r}
library(MASS)
isoMDS(dist(dissim))
```

Sal's example (this takes several minutes)
```{r}
library(vegan)
ptm <- proc.time()
example_NMDS = metaMDS(dissim, k=2) 
plot(example_NMDS)
proc.time() - ptm
```
2)
nMDS practice, same as Sal's example (skip)
```{r}
library(vegan)

#mydata <- read.table("mydata.txt", header=TRUE, row.names=1, sep=",") #using dissmim 
#mydata.mds <- metaMDS(mydata)
dissim.mds<-metaMDS(dissim) #this uses Bray distance 
# the default MDS ordination
```
Same example, but with manhattan distance 
```{r}
ptm <- proc.time()
dissim.mds<-metaMDS(dissim,distance="manhattan") 
#What determines the # of dimensions if we don't set k? In this example, ndim=2 
proc.time() - ptm
``` 
Same example, using maxit 
```{r}

dissim.mds<-metaMDS(dissim,distance="manhattan",maxit=25,trymax=25) 
```

```{r}
library(vegan)
ptm <- proc.time()
dissim.mds<-metaMDS(dissim,distance="manhattan",maxit=100,trymax=100) 
proc.time() - ptm
```











Another example, k=3 trymax=50 (skip)
```{r}
#mydata.mds.ALT <- metaMDS(mydata, distance="euclidean", k=3, trymax=50, autotransform=FALSE)
#Should we use Euclidian, Manhattan, or Bray? 

###dissim.mds.ALT<-metaMDS(dissim, distance="manhattan", k=3, trymax=50, autotransform=FALSE)

# Shows how an MDS could be performed on non-ecological data,
# where a euclidean distance metric would be appropriate. The
# transformations appropriate for ecological data are also
# turned off, so one would need to make any necessary
# transformations prior to calling the metaMDS function. This
# MDS will be 3-dimensional (k=3), and will use 50 starts from
# random configurations to avoid local minima.

#Since our data is ecological, would we want to keep the autotransform? 

dissim.mds.ALT1<-metaMDS(dissim, distance="manhattan", k=3, trymax=50) 
```
For loop that tests different k values 
```{r}
library(vegan)
dissim.mds.loop<-metaMDS(dissim, distance="manhattan", k=1, trymax=5)
  plot(dissim.mds.loop$ndim, dissim.mds.loop$stress)
for(k in 2:3)
{
  dissim.mds.loop<-metaMDS(dissim, distance="manhattan", k, trymax=5)
  plot(dissim.mds.loop$ndim, dissim.mds.loop$stress)
}
```



3) View items in the list produced by metaMDS.
```{r}
names(dissim.mds)
# mydata.mds$points: sample scores
# mydata.mds$dims: number of MDS axes or dimensions
# mydata.mds$stress: stress value of final solution
# mydata.mds$data: what was ordinated, including any
# transformations
# mydata.mds$distance: distance metric used
# mydata.mds$converged: whether solution converged or
# not (T/F)
# mydata.mds$tries: number of random initial configurations
# tried
# mydata.mds$species: scores of variables (species / taxa
# in ecology)
# mydata.mds$call: restates how the function was called
```
4) View the results of the MDS, which will display several elements of the list detailed above,
including how the function was called (call), the data set and any transformations used (data),
the distance measure (distance), the number of dimensions (dims), the final stress value
(stress), whether any convergent solution was achieved (converged), how many random initial
configurations were tried (tries), plus whether scores have been scaled, centered, or rotated.
```{r}
dissim.mds

```
5) Extract sample and variable scores. The column numbers correspond to the MDS axes, so
this will return as many columns as was specified with the k parameter in the call to
metaMDS.
```{r}
#will this code work with dissim?
variableScores <- dissim.mds$species #NA
sampleScores <- dissim.mds$points
```
6) Plot sample and variable scores in same space. Open black circles correspond to samples
and red crosses indicate taxa.
```{r}
plot(dissim.mds)
```
7) MDS plots can be customized by selecting either sites or species. Also, labels may be displayed instead of symbols by specifying type=”t”.
```{r}
 plot(dissim.mds, type="t", display=c("species"))
# text labels instead of symbols
# Crowding of text labels can be alleviated by plotting to
# a larger window

#Error in order(x) : argument 1 is not a vector
```
8) MDS plots can be further customized. By specifying type =”n”, no sample scores or variable scores will be plotted. These can then be plotted with the points() and text() commands.
For crowded plots, congestion of points and labels can be alleviated by plotting to a larger
window and by using cex to reduce the size of symbols and text.
```{r}

```

